---
title: "Presidential First Term Analysis of Salaries"
date: '`r format(Sys.Date(), "%m-%d-%Y")`'
output: 
  github_document:
    html_preview: FALSE
  pdf_document:
    toc: false
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Comparing Median Salary Among First-term Presidential Staff

Recently the Trump Administration released the first report of staff salaries.
I thought it would be interesting to do an analysis that compares it to past 
EOP offices.

For this analysis, we are going to look at the first year of Presidents 
Trump, Obama and Bush.

```{r firstTerm, echo= FALSE}
salaries %>% 
  group_by(president) %>% 
  filter(grepl("2001|2009|2017", year)) %>% 
  summarize(salary = median(salary)) %>% 
  knitr::kable()
```

`r ifelse(median(salaries$salary[salaries$year == 2017]) > median(salaries$salary[salaries$year == 2009]), paste("Here we see that Trump's median salary of", median(salaries$salary[salaries$year == 2017]), "is greater than both former president Bush and Obama.", sep = " "), aste("Here we see that Trump's median salary of", median(salaries$salary[salaries$year == 2017]), "is less than both former president Bush and Obama.", sep = " "))`

That could be expected as wages should increase with cost of living over time.

The next question I have is: How many Trump staffers are paid the maximum
compared to previous administrations? Let's take a look:

```{r highestPaidEmployees, echo = FALSE}
salaries %>% 
  filter(status == 'employee',
         grepl("2001|2009|2017", year)) %>% 
  group_by(president, salary) %>%
  summarize(count = n()) %>% 
  arrange(desc(salary)) %>% 
  top_n(1, salary) %>% 
  knitr::kable()
```

In this piece of analysis, I have excluded detailees, as their salaries are 
paid by other departments. Here I've found that Trump ties Obama for having 
the largest count of highest paid employees on staff during the first year
of their first term. Let's take another look visually to confirm what we're
seeing:

```{r payBoxplot, echo = FALSE}
salaries %>% 
  filter(status == 'employee',
         grepl("2001|2009|2017", year)) %>%
  group_by(president) %>% 
  ggplot(aes(president, salary)) +
  geom_boxplot(aes(fill = president)) +
  ggtitle('First Year, First Term Salaries By President')
```

## Staff Sizes

Let's take a look at staff sizes and salaries over time:

```{r staffSize, echo = FALSE}
salaries %>% 
  filter(grepl("2001|2009|2017", year)) %>% 
  filter(!is.na(salary)) %>% 
  group_by(year, president, party, term) %>% 
  summarize(medianSalary = median(salary),
            meanSalary = mean(salary),
            staffSize = n()) %>% 
  knitr::kable()
```

Here it appears that President Trump is spending more per staff over all,
but with a smaller staff size. This could be problematic for the president
in the long run.


So who are these top paid staff members of Trump's administration?

```{r staff, echo = FALSE}
salaries %>% 
  filter(year == 2017) %>% 
  arrange(desc(salary)) %>% 
  top_n(10, salary) %>% 
  select(employee_name, position, salary, status) %>% 
  knitr::kable()
```

Other than Mark House who is a detailee, all are employees. Probably one
of the more well known employees is Steve Bannnon, who is listed as 
Assistant to the President and Chief Strategist and Senior Counselor.

As it's noted in Wikipedia, each president gets to choose their own cabinet,
which means they can also give out their own titles as well. That makes an
analysis more difficult, so let's look and see if other presidents retain
a senior counsel on staff.

```{r seniorCounsel, echo = FALSE}
salaries %>% 
  filter(grepl('senior counsel', position)) %>% 
  select(- party, -term, -gender) %>% 
  arrange(desc(year)) %>% 
  knitr::kable()
```

Also of note, Assistants, Special Assistants and Deputy Assistants all
indicate different roles within office tiers.